maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-07-27 docs/paper/paper.tex
Fix problematic links?

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survery}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we found that
existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) lack support for transforming between multiple
representations and provide limited control over the maze generation
process.

@@

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}) but with many more options, detailed
  \secref{sec:tokenized-output-formats}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in \secref{sec:implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

\leavevmode\vadjust pre{\hypertarget{ref-pfau2024dotbydot}{}}%
Pfau, J., Merrill, W., \& Bowman, S. R. (2024). Let's think dot by dot:
Hidden computation in transformer language models. \emph{arXiv Preprint
arXiv:2404.15758}.

\leavevmode\vadjust pre{\hypertarget{ref-interpretability-survery}{}}%
Räuker, T., Ho, A., Casper, S., \& Hadfield-Menell, D. (2023). Toward
transparent ai: A survey on interpreting the inner structures of deep
neural networks. \emph{2023 IEEE Conference on Secure and Trustworthy
Machine Learning (SaTML)}, 464--483.

>>>>

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survey}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we found that
existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) lack support for transforming between multiple
representations and provide limited control over the maze generation
process.

@@

  \begin{itemize}
  \tightlist
  \item
    \texorpdfstring{\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}}}{RasterizedMazeDataset},
    utilizing
    \texorpdfstring{\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html#LatticeMaze.as_pixels}{\texttt{as\_pixels()}}}{as\_pixels()},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \texorpdfstring{\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}}{as\_ascii()}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \texorpdfstring{\href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html#MazePlot}{\texttt{MazePlot}}}{MazePlot}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \texorpdfstring{\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}}{SolvedMaze(...).as\_tokens()}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}) but with many more options, detailed
  \secref{sec:tokenized-output-formats}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in \secref{sec:implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

\leavevmode\vadjust pre{\hypertarget{ref-pfau2024dotbydot}{}}%
Pfau, J., Merrill, W., \& Bowman, S. R. (2024). Let's think dot by dot:
Hidden computation in transformer language models. \emph{arXiv Preprint
arXiv:2404.15758}. \url{https://doi.org/10.48550/arXiv.2404.15758}

\leavevmode\vadjust pre{\hypertarget{ref-interpretability-survey}{}}%
Räuker, T., Ho, A., Casper, S., \& Hadfield-Menell, D. (2023). Toward
transparent ai: A survey on interpreting the inner structures of deep
neural networks. \emph{2023 IEEE Conference on Secure and Trustworthy
Machine Learning (SaTML)}, 464--483.
\url{https://doi.org/10.1109/satml54575.2023.00039}
